Surrogate model of aligned-spin binary neutron star waveforms using Gaussian process regression

### 2017-07-27 paper/paper.tex
More reorganization/suggestions for the EOB section.

\def\l{\ell}
\def\blambda{{\bm \lambda}}
\def\bLambda{{\bm \Lambda}}
\def\blam{\bar{\lambda}}
\def\Lam{\Lambda}
\def\Lamt{\tilde{\Lambda}}
\def\dLamt{\tilde{\delta{\Lambda}}}
\def\bmu{{\bm \mu}}
\def\btheta{{\bm \theta}}

@@

Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters. The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as corrections for dynamical tides, that are accurate enough to avoid significant biases in the measured parameters. However, current implementations take at least 10 minutes to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations. We construct a frequency-domain surrogate model that requires less than 0.2~s to evaluate for a waveform with a starting frequency of 10~Hz (0.05~s for a starting frequency of 30~Hz). We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter. We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with approximately 500 waveforms instead of the $>10^5$ waveforms that would be required with standard, grid-based interpolation methods. For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.

@@

\red{Change arrow vectors to bold vectors. ${\bm x}^i$ (bold) for example $i$. $x_j$ (not bold) for coordinate/parameter/feature $j$. 
$x_j^i$ (not bold) for feature $j$ of example $i$.}

@@

%The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO), Virgo, and KAGRA which are now making routine detections of binary black hole (BBH) systems or will begin in the next few years. Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]]. From the nondetection of BNS events in aLIGO's first observing run (O1) the 90\% upper limit is still consistent with the upper end of this range [[O1UpperLimits]]. However, continued nondetection will begin to narrow down this range.

The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO) [[cite]], now making routine detections of binary black hole (BBH) systems [[cite]], as well as Virgo [[cite]], KAGRA [[cite]], and LIGO-India [[cite]]. Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]], and the nondetection of BNS events in aLIGO's first observing run (O1) is still consistent with the upper end of this range to 90\% confidence [[O1UpperLimits]]. However, continued nondetection in runs now underway will narrow down this range.

%The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters. These parameters are measured with Bayesian parameter estimation that compares the GW data with waveform templates generated on-the-fly, and require highly accurate waveform models to reliably measure the parameters with minimal systematic errors. [[Review of systematic error BNS papers...]] Previous work on estimating the tidal parameters, for example, has shown that even phase uncertainties of several radians, as is present in current 3.5-post-Newtonian waveform models, leads to systematic errors that are as large as the statistical errors [[cite]] and this also biases the recovered NS equation of state as well [[cite]]. Uncertainties in the PN waveform model also lead to significant biases in the recovered masses and spins as well [[cite]]. [[What about sky position? Does it matter?]]

The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters. These parameters, measured with Bayesian parameter estimation techniques, require highly accurate waveform models to avoid systematic errors. For example, post-Newtonian (PN) waveform models which are truncated at the currently known 3.5PN order, lead to a uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz. This is approximately the same size as the contribution to the phase from tidal effects, and as a result the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]], and this also biases the recovered NS equation of state as well [[cite]]. Uncertainties in the PN waveform model also lead to significant biases in the recovered masses and spins as well [[cite]].

The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms. Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveforms uncertainties of $\sim 10$~rad. NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger. Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]]. Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable. We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\mathbf{x})$ given its waveform parameters $\mathbf{x}$. The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce the original training set of waveforms to some desired accuracy. This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions. In previous work, this reduced basis has been constructed either with singular value decomposition (SVD) [[Michael's papers]] or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization [[All the other papers]].

Once this reduced basis is built, one can interpolate each of the coefficients $c_i(\mathbf{x})$ as a function of the waveform parameters $\mathbf{x}$. This method has been used for frequency-domain surrogates of aligned spin BBH waveforms [[Michael's papers]]. Alternatively, one can use the empirical interpolation method [[cite]] which re-expresses the linear combination of orthonormal basis functions in terms of a sum of empirical interpolating functions $B_j(t)$ times the value of the waveform $h(T_j; \mathbf{x})$ at empirical nodes $T_j$. One then interpolates the waveform $h(T_j; \mathbf{x})$ as a function of $\mathbf{x}$ at each node $T_j$. This method has been used for time domain-waveforms for nonspinning [[cite]], aligned-spin [[cite]], and precessing [[cite]] BBH waveforms as well as nonspinning BNS waveforms with tidal interactions [[cite]], as well as for frequency-domain waveforms [[cite]].

Previous work on surrogate models has focused mainly on parameter spaces $\mathbf{x}$ with $<=3$~dimensions. This allows one to approximate the coefficients $c_i(\mathbf{x})$ or waveform values $h(T_j; \mathbf{x})$ as functions of $\mathbf{x}$ using simple fitting functions [[cite]], or standard interpolation techniques such as tensor spline [[Michael, others]] or Chebyshev interpolation [[Ben]]. These interpolation techniques typically require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the 5-dimensional problem here, this is somewhat unfeasable. Recent work on optimally choosing waveforms for both EOB [[cite GPRpaper]] and NR [[cite CaltechPrecessing]] waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

@@

In this paper, we will develop methods to construct a surrogate model for the inspiral through merger of an aligned spin BNS waveform with tidal interactions using the EOB tidal description described in Hinderer et al.~[[cite1, cite2]]. This work is a continuation of the surrogate developed in [[cite]] for a nonspinning BNS waveform a using slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~[[cite, ...]]. Whereas in [[Ben]] we constructed a time-domain surrogate for the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation.

We organize the paper as follows. In Section II we describe the EOB waveform model used for constructing the surrogate and the approximations we use to reduce the dimensionality of the parameter space. In section III, we describe the details of building the surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in section IV. Then, finally discuss future improvements and applications in Section V.

\textit{Conventions:} Unless explicitly stated, we use units where $G=c=1$. To agree with the machine learning literature, we will refer to the waveform parameters (features) as $\mathbf{x}$ and the hyperparameters in the Gaussian Process Regression model as $\mathbf{\theta}$.

@@

@@

Ever since its inception~\cite{Buonanno:1998gg}, the effective-one-body (EOB) approach to the general-relativistic 2-body problem has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of spinning, nonprecessing binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter---typically binary neutron stars (BNSs) or neutron-star/black hole binaries (NSBHs). The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion---be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of such mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.

Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime.

In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below. Let $A$ and $B$ be the body labels. Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.  Explicitly, including quadrupolar and octupolar tidal effects, we have
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
\end{align}
\end{widetext}
where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star. Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$.

Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
\begin{equation}
h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{insp-plunge, pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
\end{equation} 
where $h_{\ell m}^{\textrm{insp-plunge, pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs. (A14)-(A17) of Ref.~\cite{Damour:2012yf}. Following Ref.~\cite{Dietrich:2017feu}, in the computation of $h_{22}^{\textrm{tidal}}$ we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}.

Let us now discuss the construction of the full waveform. The numerical integration starts from quasicircular initial conditions that are built following the prescription of Ref.~\cite{Buonanno:2005xu}.

The inspiral-plunge orbital dynamics is obtained by numerically integrating Hamilton's equations for the EOB Hamiltonian supplemented by a radiation-reaction force that is proportional to the GW flux derived from the $h_{\ell m}^{\textrm{insp-plunge}}$'s: see Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}.

The inspiral-plunge GW emission is obtained by evaluating Eq.~(\ref{hlminspplunge}) along the orbital dynamics up to $t=t_{\textrm{match}}^{22}$. Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when $|h_{22}^{\textrm{insp-plunge}}|$ is maximum. Let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum. Then we put
\begin{equation}
t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.
\end{equation}
In this work we only consider the dominant mode, $(\ell,m)=(2,2)$.

>>>>

%\def\l{\ell}
%\def\blambda{{\bm \lambda}}
%\def\bLambda{{\bm \Lambda}}
%\def\blam{\bar{\lambda}}
%\def\Lam{\Lambda}
%\def\Lamt{\tilde{\Lambda}}
%\def\dLamt{\tilde{\delta{\Lambda}}}
%\def\bmu{{\bm \mu}}

\def\bx{\mathbf{x}}
\def\btheta{\boldsymbol{\theta}}

@@

Fast and accurate waveform models are necessary for measuring properties of inspiraling binary neutron star systems such as the masses, spins, and tidal parameters. The effective one body (EOB) formalism has been used to produce BNS waveforms that include aligned-spin, quadrupolar and octopolar adiabatic tides, as well as corrections for dynamical tides, that are accurate enough to avoid significant biases in the measured parameters. However, current implementations take at least 10 minutes to evaluate, and this is too slow for parameter estimation codes that require $10^7$--$10^8$ sequential waveform evaluations. We construct a frequency-domain surrogate model that requires less than 0.2~s to evaluate for a waveform with a starting frequency of 10~Hz (0.05~s for a starting frequency of 30~Hz). We first reduce the 12-dimensional intrinsic parameter space of this model to the 5-dimensional parameter space that includes mass ratio, aligned-spins, and quadrupolar tides using universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameter. We then use Gaussian process regression and active learning to optimally sample the parameter space and construct a surrogate with approximately 500 waveforms instead of the $\gtrsim 10^5$ waveforms that would be required with standard, grid-based interpolation methods. For the parameter most sensitive to waveform errors, the tidal parameter, we find that the error from the resulting surrogate leads to a systematic error in the measured tidal parameter $\tilde\Lambda$ of less than $\sim 50$ which is less than 10\% of typical tidal parameters.

@@

\red{[[Change arrow vectors to bold vectors. ${\bm x}^i$ (bold) for example $i$. $x_j$ (not bold) for coordinate/parameter/feature $j$. 
$x_j^i$ (not bold) for feature $j$ of example $i$.]]}

@@

The inspiral and merger of binary neutron star (BNS) systems is one of the primary targets for second generation gravitational-wave (GW) detectors such as Advanced LIGO (aLIGO) [[cite]] now making routine detections of binary black hole (BBH) systems [[cite]], as well as Virgo [[cite]], KAGRA [[cite]], and LIGO-India [[cite]]. Available rates estimates indicate that there should be 0.4--400 events per year when aLIGO reaches its design sensitivity [[cite]], and the nondetection of BNS events in aLIGO's first observing run (O1) is still consistent with the upper end of this range to 90\% confidence [[O1UpperLimits]]. However, continued nondetection in runs now underway will decrease the upper limit of this range.

The primary quantities of interest for BNS merger observations are the sky position, masses, spins, and tidal parameters. These parameters, measured with Bayesian parameter estimation techniques, require highly accurate waveform models to avoid systematic errors. For example, post-Newtonian (PN) waveform models, which are truncated at the currently known 3.5PN order, lead to uncertainties in the waveform phase of approximately 10~rad up to a merger frequency of $\sim 1000$~Hz [[cite]]. This is approximately the same size as the contribution to the phase from tidal effects, and as a result, the systematic errors in the measured tidal parameter are as large as the tidal parameter itself [[cite]]. This also biases the recovered NS equation of state and radius [[cite]], as well as the masses and spins [[cite]].

The effective one body (EOB) formalism, on the other hand, has significantly smaller uncertainties than PN waveforms. Variations within the EOB formalism lead to uncertainties in the waveform phase of $\sim 1$~rad [[cite or say as shown below]] as compared to PN waveform uncertainties of $\sim 10$~rad. Furthermore, NR simulations of BNS systems now claim phase errors of less than 1~rad during the last $\sim 10$ orbits through merger [[cite]] and show good agreement with EOB models [[cite]]. Unfortunately, available implementations for generating EOB waveforms for BNS systems take anywhere from 10 minutes to a day to evaluate for a single waveform with a starting GW frequency of 10Hz [[cite]]. Bayesian parameter estimation algorithms such as Markov-chain Monte Carlo and nested sampling, however, typically require $10^7$--$10^8$ sequential waveform evaluations, making current EOB implementations unusable. Significant work has been done to optimize existing implementations of EOB for BBH systems [[EtienneMcWilliams]]. However, it is not clear that these optimizations can be readily adapted for the BNS model discussed here with dynamical tides. We therefore develop a surrogate model that accelerates the evaluation of EOB waveforms by interpolating between a sparsely sampled training set of waveforms within the parameter space.

Surrogate modeling techniques have had significant success in gravitational-wave data analysis for accelerating the evaluation of a waveform $h(t;\bx)$ given its waveform parameters $\bx$. The standard technique is to construct a training set of waveforms that sample the parameter space, then compress this training set by building a reduced, orthonormal basis whose linear combination can reproduce any waveform in the parameter space to some desired accuracy. This serves to interpolate the waveform as a function of time (or frequency for frequency-domain waveforms) in terms of a reduced set of coefficients corresponding to the reduced set of basis functions. In previous work, this reduced basis has been constructed either with singular value decomposition (SVD) [[Michael's papers]] or with a greedy method that chooses a subset of the training-set waveforms that are as orthogonal to each other as possible, then orthonormalizes this subset with Graham-Schmidt orthogonalization [[All the other papers]].

Once this reduced basis is built, one can interpolate each of the coefficients $c_i(\bx)$ as a function of the waveform parameters $\bx$. This method has been used for frequency-domain surrogates of aligned spin BBH waveforms [[Michael's papers]]. Alternatively, one can use the empirical interpolation method (EIM) [[cite]] which re-expresses the linear combination of orthonormal basis functions in terms of a sum of empirical interpolating functions $B_j(t)$ times the value of the waveform $h(T_j; \bx)$ at empirical nodes $T_j$. One then interpolates the waveform $h(T_j; \bx)$ as a function of $\bx$ at each node $T_j$. This method has been used for time domain-waveforms for nonspinning [[cite]], aligned-spin [[cite]], and precessing [[cite]] BBH waveforms as well as nonspinning BNS waveforms with tidal interactions [[cite]], and frequency-domain, precessing waveforms [[cite]].

Previous work on surrogate models has focused mainly on parameter spaces $\bx$ with $\le 3$~dimensions. This allows one to approximate the coefficients $c_i(\bx)$ or waveform values $h(T_j; \bx)$ as functions of $\bx$ using simple fitting functions [[cite]] or standard interpolation techniques such as tensor spline [[Michael, others]] or Chebyshev interpolation [[Ben]]. These interpolation techniques typically require waveforms evaluated on a rectangular grid, and this suffers from the curse of dimensionality; for $N$ points per dimension $d$,  $N^d$ waveforms are needed. For the 5-dimensional problem here, this is somewhat unfeasable. Recent work on optimally choosing waveforms for both EOB [[cite GPRpaper]] and NR [[cite CaltechPrecessing]] waveforms has shown that there are sufficiently accurate alternatives to grid-based interpolation that do not suffer from the curse of dimensionality.

@@

In this paper, we will develop methods to construct a surrogate model for the inspiral through merger of an aligned spin BNS waveform with tidal interactions using the EOB tidal description described in Hinderer et al.~[[cite1, cite2]]. This work is a continuation of the surrogate developed in [[cite]] for a nonspinning BNS waveform using a slightly different description for the point-particle and tidal pieces by Bernuzzi et al.~[[cite, ...]]. Whereas in [[Ben]] we constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model which significantly speeds up parameter estimation by not requiring an online Fourier transform for each waveform evaluation. It also enables additional techniques such as reduced order quadrature [[cite]] to accelerate likelihood evaluation.

We organize the paper as follows. In Section II we provide an overview of the EOB waveform model used for constructing the surrogate and the approximations we use to reduce the dimensionality of the parameter space. In section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in section IV. Then, finally discuss future improvements and applications in Section V.

\textit{Conventions:} Unless explicitly stated, we use units where $G=c=1$. To agree with the GPR literature, we will refer to the waveform parameters as $\bx$ and the hyperparameters in the GPR model as $\btheta$.

\red{[[In the introduction you should emphasize the parts of this paper that are new to the GW field: Applying GPR to a high dimensional parameter space, showing that you can choose several new training set samples then calculate them in parallel without reconstructing the reduced basis. Building a surrogate for the residual to reduce the required interpolation accuracy. The first BNS waveform that realistically tapers to zero amplitude. Active learning method that readily extends to higher dimensions. By attaching to TaylorF2, the surrogate is valid for all frequencies.]]}

\red{[[We aim to construct a sufficiently accurate waveform surrogate using as few waveform evaluations as possible, to show how one could use NR simulations/go to higher dimensions.]]}

@@

\red{[[Say somewhere that this model is based off of what is known as SEOBNRv4 (and/or 2?) in lalsuite.]]}

\red{[[The inspiral part is exactly as described in previous papers. The post-merger part is new to this paper. If this is true, we should write that somewhere.]]}

The effective-one-body (EOB) approach to the general-relativistic 2-body problem, first described in Ref.~\cite{Buonanno:1998gg}, has proven successful in modeling the dynamics and GW emission of binaries of compact objects. State-of-the-art aligned-spin EOB models~\cite{Bohe:2016gbl,Nagar:2017jdw} can accurately match hundreds of numerical-relativity (NR) simulations of aligned-spin binary black holes (BBHs) for mass ratios up to 8 and spin magnitudes up to 0.85 for unequal-mass (up to almost extremal \red{[[quantify this]]} for equal-mass) binaries. The EOB framework can also accommodate precessing-spin BBHs, showing good agreement to mildly precessing NR simulations~\cite{Babak:2016tgq}. This research program has been extended to accommodate tidal effects~\cite{Damour:2009wj,Vines:2010ca,Damour:2012yf,Bini:2012gu,Bernuzzi:2014owa,Hinderer:2016eia,Steinhoff:2016rfi,Dietrich:2017feu}, which become especially relevant in the strong-field regime for 2-body systems that contain matter---BNS or neutron-star-black-hole (NSBH) binaries. The main effect of tides is to make the gravitational interaction more attractive with respect to the vacuum case. In particular, Refs.~\cite{Hinderer:2016eia,Steinhoff:2016rfi} built upon the aligned-spin BBH EOB model of Ref.~\cite{Taracchini:2013rva} and proposed a way to include the effect of dynamical tides. Neutron stars that are part of a compact-object binary will deform in the tidal field generated by the companion---be it a black hole or another neutron star. Because of binary motion, the forcing tidal field varies at the orbital frequency. Thus, in the late stages of the inspiral, the characteristic $f$-mode frequency of the neutron star can be dynamically approached, resulting in a resonant excitation of the f-mode. The net effect is an amplification of tidal effects as compared to the adiabatic limit, which assumes that the $f$-mode frequency is much larger than the frequency of the forcing tidal field.

We note that, to date, this is the only tidal EOB model that includes the effect of spin \red{[[Sebastiano also has a model that includes spin and tides.]]}. More precisely, spin effects are those of the point-mass part of the model~\cite{Bohe:2016gbl}, that was calibrated to 141 NR simulations of aligned-spins BBHs. The model, however, does not account for the quadrupole that the neutron stars acquire when they are spinning. This is a genuine matter effect that appears at lower PN order as compared to the tidal effects that are currently modeled. The importance of this physical effect in the context of data analysis for ground-based GW interferometers is currently being studied~\cite{IanTanja}.

@@

Dynamical tidal effects are implemented in the EOB model through a modification of the radial potential $\Delta_u$, which is the $tt$-component of the metric of the effective spacetime. \red{[[Show how $\Delta_u$ enters the Hamiltonian.]]}. In this paper we adopt the tidally-augmented expression of $\Delta_u$ that is discussed in Appendix~A of Ref.~\cite{Steinhoff:2016rfi}, namely
\begin{equation}
\Delta_u = \Delta_u^{\textrm{pm}} + \Delta_u^{\textrm{DT}}\,,
\end{equation}
where $\Delta_u^{\textrm{pm}}$ is the 4PN-accurate point-mass term of the underlying BBH EOB model of Ref.~\cite{Bohe:2016gbl} (see Eq.~(2.2) therein) and $\Delta_u^{\textrm{DT}}$ is the contribution due to dynamical tides. We consider the general case of a binary that may consist of two neutron stars; if either component of the binary is a black hole, then vanishing values of the tidal polarizabilities are employed in the formulas below. Let $A$ and $B$ be the body labels. Let $M_{A,B}$ be the mass of either body and $M=M_A+M_B$ be the total mass of the binary.  Explicitly, including quadrupolar and octupolar tidal effects, we have
\begin{widetext}
\begin{align}
 \Delta_u^{\textrm{DT}} =&  - 3\,\Lambda_{2,\textrm{dyn}}^{A}(u)X_A^4 X_B\,u^6  \left[ 1 + \frac{5}{2} X_Au + \left(3 + \frac{1}{8}X_A + \frac{337}{28} X_A^2\right)u^2 \right]\nonumber\\
 &- 15\,\Lambda_{3,\textrm{dyn}}^{A}(u)X_A^6X_B\,u^8 \left[1 + \left(-2 + \frac{15}{2}X_A\right)u + \left(\frac{8}{3} - \frac{311}{24}X_A + \frac{110}{3}X_A^2\right) u^2\right] + (A \leftrightarrow B)\,,
\end{align}
\end{widetext}
where $X_{A,B}=M_{A,B}/M$, $u=1/r$ is the inverse of the ($M$-rescaled) EOB radial coordinate $r$ and $\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)$ are the dimensionless $2^{\ell}$-polar dynamical tidal polarizabilities. Within the dynamical tides model, the tidal polarizabilities are not constant, but rather depend on the orbital separation and on the values of the $f$-mode frequencies $\omega_{0\ell}^{A,B}$. In particular, the dimensionless dynamical tidal polarizability reads
\begin{equation}
\Lambda_{\ell,\textrm{dyn}}^{A,B}(u)=\Lambda_{\ell}^{A,B}\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})\,,
\end{equation}
where $\Lambda_{\ell}^{A,B}$ is the dimensionless adiabatic tidal polarizability
\begin{equation}
\Lambda_{\ell}^{A,B}=\frac{2}{(2\ell-1)!!}\frac{k^{A,B}_{\ell}}{C_{A,B}^{2\ell+1}}\,,
\end{equation}
with $k^{A,B}_{\ell}$ the tidal Love number and $C_{A,B}=R_{A,B}/M_{A,B}$ the neutron star compactness, which depends on the areal radius $R_{A,B}$ of the neutron star. Here $\hat{k}_{\ell\,\textrm{dyn}}(u;\omega_{0\ell}^{A,B})$ is the separation-dependent, dimensionless enhancement factor given in Eq.~(11) of Ref.~\cite{Dietrich:2017feu}, which depends on the value of the $f$-mode $2^\ell$-pole angular frequency, $\omega_{0\ell}^{A,B}$. \red{[[Does it need to be made clear that the $f$-mode is rescaled by the individual mass not the total mass?]]} This enhancement factor is 1 for low orbital frequencies, and increases by about \red{x\%} as $\ell$ times the orbital frequency approaches the $f$-mode frequency.

Tidal effects also enter the dissipative part of the model. In particular, the point-mass $\ell=2,3$ inspiral-plunge waveform modes are corrected by tidal terms
\begin{equation}
h_{\ell m}^{\textrm{insp-plunge}} = h_{\ell m}^{\textrm{pm}} + h_{\ell m}^{\textrm{tidal}}\,,\label{hlminspplunge}
\end{equation} 
where the point-mass piece $h_{\ell m}^{\textrm{pm}}$ is discussed in Section~II.B of Ref.~\cite{Bohe:2016gbl} and the tidal piece $h_{\ell m}^{\textrm{tidal}}$ is given by Eqs. (A14)-(A17) of Ref.~\cite{Damour:2012yf}. Following Ref.~\cite{Dietrich:2017feu} in the computation of $h_{22}^{\textrm{tidal}}$, we include a dynamical enhancement factor that depends on the orbital separation and on the $f$-mode frequency, and is given in Eq.~(15) of Ref.~\cite{Dietrich:2017feu}. The waveform modes $h_{\ell m}^{\textrm{insp-plunge}}$ are then used to calculate the GW flux (Eqs.~(5) and (6) in Ref.~\cite{Dietrich:2017feu}) from which the radiation-reaction force is derived.

Let us now discuss the construction of the full waveform. The Hamiltonian equations of motion with the above radiation-reaction force are numerically integrated beginning with the quasicircular initial conditions described in Ref.~\cite{Buonanno:2005xu}. The inspiral-plunge waveform is obtained by evaluating Eq.~(\ref{hlminspplunge}) using the solution to the orbital dynamics up to some end time $t=t_{\textrm{match}}^{22}$ that we now define. Let $t_{\textrm{peak}}^{\textrm{amp}}$ be the time when the amplitude of the $(\ell,m)=(2,2)$ mode $|h_{22}^{\textrm{insp-plunge}}|$ is maximum, and let $t_{\textrm{peak}}^{\textrm{freq}}$ be the time when the inspiral-plunge GW frequency $\omega_{22}^{\textrm{insp-plunge}}$ is maximum. We then choose the stopping condition to be the earliest of these two times
\begin{equation}
t_{\textrm{match}}^{22} = \min \left(t_{\textrm{peak}}^{\textrm{amp}},t_{\textrm{peak}}^{\textrm{freq}}\right)\,.
\end{equation}
\red{[[Explain/motivate/justify this choice in a sentence or two. Why don't you just pick one of the conditions?]]} In our final waveform model we only use the dominant $(\ell,m)=(2,2)$ mode in constructing the final waveform, although we use all modes in evaluating the radiation reaction force.
